The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2024-11-01 Paper.qmd
Paper: minor edits throughout.
Regularize: fixed mistake that led to Family conflict not being included in the model of caregiver response to last instance of child signaling.

Compared to chimpanzee mothers, mothers in hunter-gatherer societies have shorter interbirth intervals and produce offspring that require many more years of parental care [@davison2021; @davison2022]. Unlike chimpanzee offspring, human offspring in ancestral populations therefore often grew up with multiple older and/or younger siblings who were potential competitors for parental investment and yet who could also provide (or require) allocare [@helfrecht2016; @lawson2016; hagen2009].

@@

Children’s potential to invest in their families can also cause conflicts with their parents. Older children might prefer to spend their time investing in romantic or other social relationships, or learning valuable skills, whereas parents might prefer them to care for their siblings, acquire or prepare food, or complete other household chores [@lowe2003; @syme2023a]. Conflicts of interest over investment in the family also appear to extend to when a child starts having their own children [@kushnick2023; @emlen1995], with parents potentially benefiting from delaying the reproductive careers of certain offspring in order to gain more allocare for their other children [@flinn1989; @emlen1992] or speeding up a child’s departure from the household so as to provide more parental investment to other children [@kushnick2023].

@@

Throughout human evolutionary history, individuals often found themselves in situations in which they received fitness-relevant benefits from others that could not easily be replaced [@apicella2012; @apicella2019; @hill2011; @tooby1996]. In these instances, natural selection is expected to have favored those who provided support to others if 1) such help meant one could expect to receive fitness benefits from those helped in the future and 2) the costs of providing help were outweighed by the benefits to the helper on average [@andré2010; @hamilton1964; @tooby1996; @trivers1971].

@@

Crying is the most studied way children signal need. Crying is a human universal [@gracanin2018] with few suspected non-signaling functions outside of the role vocal cries might play in the development of vocalization [@wells2003] and the functions tears have in terms of cleaning, protecting, and lubricating the eye [@murube2009; @sznycer]. Hunger, pain, anger, fear, and separation are all common causes of crying [@vingerhoets2013; @wells2003], although tear production also often happens with joyful events [@sznycer; @vingerhoets2013; @zickfeld].

@@

Although widely accepted as a signal, there is no consensus on why emotional crying might elicit positive responses. As the costs and benefits of signaling and gaining investment change with age, it is likely the strategic logic of crying changes too. It could be vital for newborns to signal health and vigor to reduce the risk of neglect or infanticide experienced by children with poor survival prospects [@daly1984; @soltis2004]. As crying comes with physiological costs [@rao1997; @thureen1998], it could serve as a credible signal of vigor if these costs are too high for unhealthy infants to bear [@furlow1997; @lummaa1998; @soltis2004; @wells2003]. Alternatively, crying might impose costs on caregivers by, e.g., attracting predators or interfering with important activities, motivating caregivers to invest more to stop the crying [so-called blackmail or bargaining models; @xx, @yy]. More generally, contact calls between parents and offspring have evolved in numerous species, so cries might simply be a means for infants to elicit attention from caregivers [@mehr2021]. For older children, the relatively insignificant energetic costs of crying might not be enough to ensure it is a credible signal when there are conflicts of interest. For this reason, crying among older children might be more similar to adult crying, which can increase support and empathy but sometimes also reduces reputation and increases stigmatization [@cornelius2001; @cretser1982; @hendriks2006; @jesser1989; @vingerhoets2016]. Such costs could be a means by which crying can still be a credible signal of need if they dissuade potential fakers from making the signal [@sznycer] when there are moderate conflicts of interest.

@@

The likelihood of positive, neutral, and negative responses to the last time a child became sad or cried. A: Effect of punishment being implicated in the cause (0 = no punishment, 1 = punishment.) B: Effect of discomfort, pain, injury, or illness being implicated in the cause (0 = discomfort, pain, injury, or illness, 1 = discomfort, pain, injury, or illness.) Legend key: -1 = negative response, 0 = neutral response, 1 = positive response (all from child's perspective).
:::

@@

Caregivers reported a wide range of child signaling frequencies. About half of children were reported to cry or throw tantrums more than once a month and slightly less than 40% were reported to engage in them more than once a week. Although similarly common overall, tantrums more common than crying among older children and might be a under-appreciated way by which children might signal for increased support based on these findings and a longitudinal study of temper tantrums and otherchild behaviors which found 50% of boys 30% of girls engaged in temper tantrums at age 9 among a representative, non-clincial sample of 252 children in Berkeley, California [@@macfarlane1962].

As seen in Western cultures [@zeifman2013; @potegal2003; @koch2003; @macfarlane1962], the frequencies of crying and temper tantrums decreased with age, as did the frequency of child sadness, summed signal frequency, and signal cost. One possible reason for this is that as children mature, they are more capable of fulfilling their own desires, more able to ask others for what they want, and less vulnerable to threats. The effects of age on signaling are also likely due to how age influences the costliness of signaling need. For example, while crying may be costly among the youngest children through its energetic costs, these costs are likely insignificant among older children, who might instead face costs in the form of reputational harm from crying in some contexts but not others. In times when an older child's valuation of potential increased support is small, such costs may be prohibitive. However, in times when the child places more value on potential support, those same costs could allow for a child to credibly signal need if they are too high to be worth it for deceptive signaling.

Consistent with signaling models which emphasize the role of conflicts of interest in favoring the use of costlier signals, conflict and signaling behavior were highly correlated. Although conflicts of interest do not always lead to behavioral conflict [@altmann1980; @bateson1994], it is likely many, if not most, conflicts are ultimately caused by conflicts of interest, potentially making it a useful proxy for the underlying conflicts of interest that may increase the benefits of deception for children and the benefits of greater skepticism for parents\[\[5\]\]. If this is the case, the positive correlation between signaling effort and conflict frequency may represent children's attempts at credibly signaling their value of additional investment or the costliness of punishment [@syme2019] to parents who would have good reason to doubt the trustworthiness of simple verbal requests (or the trustworthiness of facial sadness and crying for those resorting to temper tantrums). However, it is also likely that parents find some forms of child signaling aversive, and therefore see it as conflict in and of itself.

We found evidence of an interaction between how often a child cared for their siblings and the sex of the child in our lasso models examining signal frequencies and cost, whereby there was a positive relationship between the frequency and summed costliness of child signals of need and alloparenting effort for girls but no relationship or a slightly negative one for boys. In many societies, girls do more alloparenting than boys [@helfrecht2016; @weisner1977; @barry1971; @turke1988], something which matches findings from focus groups of Utilian mothers \[add info from an analysis here?\]. If this difference was the outcome of caretakers pressuring girls to do more or resulted in the expectation of more alloparenting, the increase in signaling with alloparenting effort seen among girls may be the result of a strategy to signal the costliness of their efforts in order to invest more in relationships and status seeking behaviors outside the household.

@@

However, child alloparenting effort was negatively associated with caregiver reports of child conflict. One potential reason for this is since parents presumably value the child alloparenting within the family, they see their relationships with children who invest more in the household as less conflictual. In line with this reasoning, we found evidence of a child alloparenting effort by child age interaction, with our model predicting increased levels of conflict for older children with greater alloparenting \[if we say this we should report the results of that lasso too\]. Ideally, we could control for signaling behavior to test this. However, the bidirectional relationship between child signaling and conflict with caretakers prevents this (explain better). It is also possible that our hypothesis about conflict over investment in the household is incorrect.

There was a positive association between the frequency of child sadness and caretaker perceptions of child neediness within the family. However, the other signaling measures examined had coefficients set to 0 in the same lasso model despite the expectation that their greater costliness would make them more credible sources of information to parents. One reason for this might be that costs of sadness, while minor short-term, can be significant long-term. In line with this reasoning, depression, which often involves an exteneded period of a lack of interest in one's normal activities and decreased investment in oneself and one's cooperative relationships, has been suggested to capable of credibly signal need in times of substantial conflicts of interest [@watson2002; @hagen2003].

Another possibility is that parents are using cues from sadness (i.e., withdrawal and rumination) to evaluate the relative neediness of children. As cues are not created for the purpose of communication, they are completely reliable if perceived and interpreted correctly. Therefore, if some cues are hard to impossible to fake, they may be more reliable than most signals. That said, it is likely that some symptoms of sadness are simultaneously the result of both signals and cues, something which may have been favored by natural selection if the threat of deception meant cue-level symptoms were insufficient in eliciting support without upregulation, something which has been suggested to be the case with various illness symptoms [@tiokhin2016; @steinkopf2015].

@@

When the cause involved punishment, however, parents were more likely to respond negatively and less likely to respond positively. In these circumstances, children may be signaling about the severity of a punishment [@syme2019]. However, parents may be reluctant to reduce punishments if the behavior that warranted the punishment was repeated or severe. They may also react with scolding or further punishment. That said, our model predicts positive responses in `r signif(updated_out$estimate[3] * 100, 2)`% of cases (when all other predictors are held at their mean values), suggesting that children are successful in negotiating better outcomes at least some of the time.

Although not a significant relationship, the negative effect of the family conflict being a reported cause of signaling on the likelihood of a positive response may reflect tradeoffs within the family, whereby investment in one child cannot go to another. Caretakers may have also responded more negatively when the cause involved family conflict, if it involved a child imposing costs on other family members.

@@

Despite the differences in wealth, job certainty, discrimination, and living conditions [@davis2014; @garcia2017; @garcia2018; @stonich2000] between recent immigrants from the mainland and native Utilians, primary language, immigration status, and residence in Camponado did not predict child signaling in our lasso regressions. However, neighborhood quality, which was strongly negaltively correlated with residence in Camponado, (check others) was preserved in the lasso regressions predicting every signaling measure but sadness and was among the variables with the largest negative coefficients. Monthly income was also negatively associated with crying frequency,~~ tantrum frequency (too small of a coefficient to mention here?),~~ summed signal frequency, and signal cost albeit with smaller coefficients.

@@

>>>>

Compared to chimpanzee mothers, mothers in hunter-gatherer societies have shorter interbirth intervals and produce offspring that require many more years of parental care [@davison2021; @davison2022]. Unlike chimpanzee offspring, human offspring in ancestral populations therefore often grew up with multiple older and/or younger siblings who were potential competitors for parental investment and yet who could also provide (or require) allocare [@helfrecht2016; @lawson2016; @hagen2009].

@@

Children’s potential to invest in their families can also cause conflicts with their parents. Older children might prefer to spend their time investing in romantic or other social relationships, or learning valuable skills, whereas parents might prefer them to care for their siblings, acquire or prepare food, or complete other household chores [@lowe2003; @syme2023a]. Conflicts of interest over investment in the family also appear to extend to when a child starts having their own children [@kushnick2023; @emlen1995], with parents potentially benefiting from delaying the reproductive careers of certain offspring in order to gain more allocare for their other children [@flinn1989; @emlen1992] or speeding up a child’s departure from the household in order to provide more parental investment to other children [@kushnick2023].

@@

Saved for now: Throughout human evolutionary history, individuals often found themselves in situations in which they received fitness-relevant benefits from others that could not easily be replaced [@apicella2012; @apicella2019; @hill2011; @tooby1996]. In these instances, natural selection is expected to have favored those who provided support to others if 1) such help meant one could expect to receive fitness benefits from those helped in the future and 2) the costs of providing help were outweighed by the benefits to the helper on average [@andré2010; @hamilton1964; @tooby1996; @trivers1971].

@@

Crying is the most studied way children signal need. Crying is a human universal [@gracanin2018] with few suspected non-signaling functions outside of the role vocal cries might play in the development of vocalization [@wells2003] and the functions tears have in terms of cleaning, protecting, and lubricating the eye [@murube2009; @sznycer2023]. Hunger, pain, anger, fear, and separation are all common causes of crying [@vingerhoets2013; @wells2003], although tear production also often happens with joyful events [@sznycer2023; @vingerhoets2013; @zickfeld].

@@

Although widely accepted as a signal, there is no consensus on why emotional crying might elicit positive responses. As the costs and benefits of signaling and gaining investment change with age, it is likely the strategic logic of crying changes too. It could be vital for newborns to signal health and vigor to reduce the risk of neglect or infanticide experienced by children with poor survival prospects [@daly1984; @soltis2004]. As crying comes with physiological costs [@rao1997; @thureen1998], it could serve as a credible signal of vigor if these costs are too high for unhealthy infants to bear [@furlow1997; @lummaa1998; @soltis2004; @wells2003]. Alternatively, crying might impose costs on caregivers by, e.g., attracting predators or interfering with important activities, motivating caregivers to invest more to stop the crying [so-called blackmail or bargaining models; @xx, @yy]. More generally, contact calls between parents and offspring have evolved in numerous species, so cries might simply be a means for infants to elicit attention from caregivers [@mehr2021]. For older children, the relatively insignificant energetic costs of crying might not be enough to ensure it is a credible signal when there are conflicts of interest. For this reason, crying among older children might be more similar to adult crying, which can increase support and empathy but sometimes also reduces reputation and increases stigmatization [@cornelius2001; @cretser1982; @hendriks2006; @jesser1989; @vingerhoets2016]. Such costs could be a means by which crying can still be a credible signal of need if they dissuade potential fakers from making the signal [@sznycer2023] when there are moderate conflicts of interest.

@@

Effects of selected predictors of the likelihood of positive, neutral, and negative responses to the last time a child became sad or cried. 1 = evidence of the cause in the caregiver report; 0 = no evidence. The endpoints of the lines were dodged slightly for greater visibility. Dot size represents xxxxxxx. See @sfig-pot-caregiver-punish for the effect of the cause involving punishment.
:::

@@

Caregivers reported a wide range of child signaling frequencies. About half of children were reported to cry or throw tantrums more than once a month and slightly less than 40% were reported to engage in them more than once a week. Although similarly common overall, tantrums more common than crying among older children and might be an under-appreciated way by which children signal for increased support based on these findings and a longitudinal study of temper tantrums and other child behaviors which found 50% of boys 30% of girls engaged in temper tantrums at age 9 among a representative, non-clincial sample of 252 children in Berkeley, California [@@macfarlane1962].

As seen in Western cultures [@zeifman2013; @potegal2003; @koch2003; @macfarlane1962], the frequencies of crying and temper tantrums decreased with age, as did the frequency of child sadness, summed signal frequency, and signal cost. One possible reason for this is that as children mature, they are more capable of fulfilling their own desires, more able to ask others for what they want, and less vulnerable to threats. The effects of age on signaling are also likely due to how age influences the costliness of signaling need. For example, while crying may be costly among the youngest children through its energetic costs, these costs are likely insignificant among older children, who might instead face costs in the form of reputational harm from crying in some contexts but not others. In times when an older child's valuation of potential increased support is small, such costs may be prohibitive. However, in times when a child places more value on potential support, those same costs could allow for a child to credibly signal need if they are too high to be worth it for deceptive signaling but still made up for by the benefits for credible signalers.

Consistent with signaling models which emphasize the role of conflicts of interest in favoring the use of costlier signals, conflict and signaling behavior were highly correlated. Although conflicts of interest do not always lead to behavioral conflict [@altmann1980; @bateson1994], it is likely many, if not most, conflicts are ultimately caused by conflicts of interest, potentially making it a useful proxy for the underlying conflicts of interest that may increase the benefits of deception for children and the benefits of greater skepticism for parents\[\[5\]\]. If this is the case, the positive correlation between signaling effort and conflict frequency may represent children's attempts at credibly signaling their value of additional investment or the costliness of punishment [@syme2019] to parents who would have good reason to doubt the trustworthiness of simple verbal requests (or the trustworthiness of facial sadness and crying for those resorting to temper tantrums). It is also likely that parents find some forms of child signaling aversive, and therefore see it as conflict in and of itself.

We found evidence of an interaction between how often a child cared for their siblings and the sex of the child in our lasso models examining signal frequencies and cost, whereby there was a positive relationship between the frequency and summed costliness of child signals of need and alloparenting effort for girls but no relationship or a slightly negative one for boys. In many societies, girls do more alloparenting than boys [@helfrecht2016; @weisner1977; @barry1971; @turke1988], something which matches findings from focus groups of Utilian mothers. If this difference was the outcome of caretakers pressuring girls to do more or resulted in the expectation of more alloparenting, the increase in signaling with alloparenting effort seen among girls may be the result of a strategy to signal the costliness of their efforts in order to invest more in relationships and status seeking behaviors outside the household.

@@

However, child alloparenting effort was negatively associated with caregiver reports of child conflict. One potential reason for this is since parents presumably value the child alloparenting within the family, they see their relationships with children who invest more in the household as less conflictual. In line with this reasoning, we found evidence of a child alloparenting effort by child age interaction, with our model predicting increased levels of conflict for older children with greater alloparenting \[if we say this we should report the results of that lasso too\]. Ideally, we could control for signaling behavior to test this. However, the bidirectional relationship between child signaling and conflict with caretakers prevents this. It is also possible that our hypothesis about conflict over investment in the household is incorrect.

There was a positive association between the frequency of child sadness and caretaker perceptions of child neediness within the family. However, the other signaling measures examined had coefficients set to 0 in the same lasso model despite the expectation that their greater costliness would make them more credible sources of information to parents. One reason for this might be that costs of sadness, while minor short-term, can be significant long-term. These costs are clearly seen in depression, which often involves an extended period of a lack of interest in one's normal activities and decreased investment in oneself and one's cooperative relationships and has been suggested to capable of credibly signal need in times of substantial conflicts of interest [@watson2002; @hagen2003].

Another possibility is that parents are using cues from sadness (i.e., withdrawal and rumination) to evaluate the relative neediness of children. As cues are not created for the purpose of communication, they are completely reliable if perceived and interpreted correctly. Therefore, if some cues are hard to or are impossible to fake, they may be more reliable than most signals. That said, it is likely that some symptoms of sadness are simultaneously the result of both signals and cues, something which may have been favored by natural selection if the threat of deception meant cue-level symptoms were insufficient in eliciting support without upregulation, something which has been suggested to be the case with various illness symptoms [@tiokhin2016; @steinkopf2015].

@@

When the cause involved the child having made a transgression, punishment, or the loss of privleges or an item, however, parents were more likely to respond negatively and less likely to respond positively. In these circumstances, children may be signaling about the severity of a punishment [@syme2019]. However, parents may be reluctant to reduce punishments if the behavior that warranted the punishment was repeated or severe. They may also react with scolding or further punishment. That said, our model predicts positive responses in `r signif(updated_out$estimate[3] * 100, 2)`% of cases (when all other predictors are held at their mean values), suggesting that children are successful in negotiating better outcomes at least some of the time.

Parents were also more likely to respond negatively and less likely to respond positively to children when the cause involved family conflict. This might reflect tradeoffs within the family, whereby investment in one child cannot go to another. Caretakers may have also responded more negatively when the cause involved family conflict, if it involved a child imposing costs on other family members.

@@

Despite the differences in wealth, job certainty, discrimination, and living conditions [@davis2014; @garcia2017; @garcia2018; @stonich2000] between recent immigrants from the mainland and native Utilians, primary language, immigration status, and residence in Camponado did not predict child signaling in our lasso regressions. However, neighborhood quality, which was strongly negatively correlated with residence in Camponado, (check others) was preserved in the lasso regressions predicting every signaling measure but sadness and was among the variables with the largest negative coefficients. Monthly income was also negatively associated with crying frequency,~~ tantrum frequency (too small of a coefficient to mention here?),~~ summed signal frequency, and signal cost albeit with smaller coefficients.

@@

::: {#sfig-pot-caregiver-punish}
![](Figures/plot_caregiver_punish.svg)

The likelihood of positive, neutral, and negative responses to the last time a child became sad or cried when punishment was implicated in the cause (0 = no punishment, 1 = punishment.)
:::
